---
title: "Mini Project 4"
author: "Sophie Li, Rocio Jaime"
date: "November 15, 2017"
output: 
  html_document:
    code_folding: hide
---
#The Popularity of Sequels Over the Decades
##Using SQL and the imdb database

The release of Star Wars: The Last Jedi comes just months after the franchise's 30th anniversary. In only three decades, Star Wars expanded to a series of eight main films (with more to come), with countless spinoffs and adaptations.

If it seems like these rapid-fire releases have expanded beyond the Galactic Empire, you're not alone. In an article written for ESPN's sports and pop-culture blog, Grantland, Mark Harris claimed that " franchises are not a big part of the movie business. They are not the biggest part of the movie business. They are the movie business."
We decided to test this popular assumption by tracking the number of sequels per original movie over time. What we found confirmed Harris's claim: The frequency of sequels, remakes, and spinoffs has increased dramatically-not only are studios making sequels at a faster pace for recent movies, they're also digging into the past for inspiration for the future.
```{r, message = FALSE, warning = FALSE, fig.height=15, fig.width=20}
library(mdsr)
library(RMySQL)

db <- dbConnect_scidb(dbname = "imdb")

#Gets info on sequels, remakes, and spinoffs
#Uses indexes on type_ids
sequel_info <- db %>%
  dbGetQuery("SELECT ml.movie_id, t.title, mi.info AS original_runtime, ml.linked_movie_id AS sequel_id, t2.title AS sequel_title, mi2.info AS sequel_runtime, ml.link_type_id, t.production_year AS original_year, t2.production_year AS sequel_year
FROM movie_link ml
JOIN title t ON t.id = ml.movie_id
JOIN title t2 ON t2.id = linked_movie_id
JOIN movie_info mi ON mi.movie_id = ml.movie_id
JOIN movie_info mi2 ON mi2.movie_id = ml.linked_movie_id
WHERE ml.link_type_id IN (2, 4, 12)
	AND t.kind_id = 1
  AND mi.info_type_id = 1 
  AND mi2.info_type_id = 1
  AND mi.info > 89
  AND mi2.info > 89;
")

all_movies <- db %>%
  dbGetQuery("SELECT t.id AS movie_id, t.title, mi.info AS original_runtime, t.production_year AS original_year
FROM title t
JOIN movie_info mi ON mi.movie_id = t.id
WHERE t.kind_id = 1
  AND mi.info_type_id = 1 
  AND mi.info > 89
")


#gets gross of movies (max cumulative gross in USA only, due to constraints w/ format of data)
gross <- db %>% 
  dbGetQuery('SELECT mi.movie_id, mi.info AS gross
             FROM movie_info mi
             WHERE mi.info_type_id = 107;
             ') %>% 
  filter(grepl('\\$', gross)) %>% 
  filter(grepl('\\(USA)', gross)) %>% 
  mutate(gross = gsub('\\s.*', '', gross)) %>%
  mutate(gross = gsub('\\D', '', gross)) %>%
  mutate(gross = as.numeric(gross)) %>%
  group_by(movie_id) %>% 
  summarize(max_gross = max(gross)) %>% 
  arrange(desc(max_gross))

#filters for duplicates (so only original movie shows up, w/ sequels linking to it)
#adds decade column
sequel_info <- sequel_info %>% 
  filter(!(title %in% sequel_title)) %>% 
  mutate(decade = 10 * floor(original_year / 10)) %>% 
  select(movie_id, title, original_year, decade, sequel_title, sequel_year, link_type_id) %>%
  left_join(gross, movie_id = movie_id) %>% 
  unique()

all_movies <- all_movies %>% 
  mutate(decade = 10 * floor(original_year / 10)) %>% 
  left_join(gross, movie_id = movie_id) %>% 
  unique()

#Function that finds top 10 grossing movies
top10 <- function(data) {
  data %>%
    group_by(title) %>%
    summarize(top_movies = max(max_gross)) %>%
    arrange(desc(top_movies)) %>%
    head(10)
}

#Finds top 10 grossing movies per decade
top_sequels_decade <- sequel_info %>% 
  filter(!is.na(max_gross)) %>% 
  group_by(decade) %>%
  do(top10(.))

#Calculates popular sequels
pop_seq <- sequel_info %>% 
  filter(title %in% top_sequels_decade$title & !is.na(max_gross)) %>% 
  select(decade, title, original_year, sequel_title, sequel_year, link_type_id) %>% 
  mutate(title = reorder(title, original_year)) %>% 
  mutate(link_type_id = factor(link_type_id))

ggplot(pop_seq, aes(x = original_year, y = title)) +
  geom_segment(aes(xend = sequel_year, yend = title), size = 2) +
  geom_point(size = 3) +
  geom_point(aes(x = sequel_year, y = title, color = link_type_id), size = 3) +
  labs(color = 'Type of Movie', title = "Sequels/Remakes/Spinoffs Produced by Decade", x = "Release Year of Original Movie", y = "Movie Title") +
  theme(axis.text.x = element_text(size= 15),
        axis.text.y = element_text(size = 11),
        plot.title = element_text(size = 28),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 16),
        legend.text = element_text(size= 14),
        axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size = 20)) +
  scale_color_discrete(labels = c('Sequel', 'Remake', 'Spin-off')) +
  scale_y_discrete(labels = scales::wrap_format(40)) +
  scale_x_continuous(limits = c(1910, 2020), expand = c(0, 0)) +
  facet_wrap( ~ decade, nrow = 5, scales = "free_y") 
```
Our analyses were made by searching through the IMDb (Internet Movie Database), and creating a list of every US movie greater than ninety minutes that had either a sequel, remake, or spinoff. We then grouped each movie by the decade of the original movie's release date and only included the top ten grossing movies of that decade. The movies were then graphed and separated by their decade group in order to more clearly see the progression of sequel release patterns as time passed.

There seem to be two main trends occurring over the years. First, that successful movie franchises are growing larger in terms of number of sequels per top 10 movie. This may be caused by behaviors from both moviegoers and movie executives. According to Martin Kaplan, a professor of film and media at the University of Southern California, "in an unsettling world people may gravitate toward the familiar...sequels are a kind of comfort food."
This trend occurs on the production side as well.  Because of this constant baseline audience, movie studios may be less likely to spend money developing risky, original movies.

The renowned film critic Roger Ebert claimed "no movie executive has ever been fired for greenlighting a sequel. Once a brand has been established in the marketplace, it makes sound business sense to repeat the formula." With both sides of the producing and consuming cycles supporting the creation and consumption of sequels, it's no wonder they're as popular to make and watch as they are.
  
The second trend that we noticed was that the period of time between the first movie of the series and the last movie of the series has grown over the years.This trend seems logical, given that older movies have more time to be remade-but it still shows that Hollywood studios turn to the past for new inspiration. Though the sequels, remakes, and reboots that come from this may have little in common with their predecessors, they still contribute to the perception that Hollywood is losing its originality.

The third trend was the most striking -when we compared the decade groups, we saw that the time between movies and their sequels had decreased significantly. This, in combination with the two other trends, heavily influence public opinion towards movie franchises and sequels in general. Decreased time between movies, along with increased number of movies and longevity, contribute to moviegoers' fatigue and critics frustration-it's hard to appreciate an individual movie on its own merits when it's linked to several others.

Of course, sequels in and of themselves are not necessarily bad per se, as they date back to ancient Greece with famous playwrights such as Sophocles writing follow ups to their plays, including the Oedipus trilogy. When thoughtfully crafted, sequels can prove themselves to be exciting, great successes, with examples including Godfather: Part II and Terminator 2: Judgement Day. But when sequels are being produced at their current rate, rapidly and following nearly every popular film, it feels as though producers are sacrificing quality and creativity in order to squeeze out as much profit from a franchise as possible. This leaves us with movies like Hangover: Part III.

Aaron Hatch from The Artifice takes an opposing stance: that, in proportional terms, sequels are actually less common. And he's right.
```{r, warning= FALSE}
nseq <- sequel_info %>% 
  filter(!is.na(max_gross)) %>% 
  group_by(decade, title) %>% 
  summarize(n = n()) %>% 
  group_by(decade) %>% 
  summarize(numSeq = sum(n))

nall <- all_movies %>% 
  filter(!is.na(max_gross)) %>% 
  group_by(decade, title) %>% 
  summarize(n = n()) %>% 
  group_by(decade) %>% 
  summarize(numAll = sum(n))
nratio <- nall %>% 
  left_join(nseq) %>% 
  mutate(ratio = numSeq / numAll)

ggplot(nratio, aes(x = decade, y = ratio)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(from = 1920, to = 2020, by = 10), limits = c(1920, 2010)) + 
  labs(title = 'Ratio of Sequels to Total Movies Produced in a Decade', y = 'Ratio', x = 'Decade')
```

The figure above is a line graph of sequels, remakes, and spinoffs as a proportion of all movies produced. If we only used this as criteria, claims about the proportions of sequels would be unfounded.

However, this approach is problematic because it disregards all other factors that influence human opinion. For example, the mean length between movies decreases by a far larger factor than the proportion of sequels to original movies:

```{r}
sequel_info %>% 
  filter(!is.na(max_gross)) %>% 
  group_by(decade, title) %>% 
  arrange(sequel_year, .by_group = TRUE) %>% 
  summarize(mPerMovie = mean(diff(c(original_year, sequel_year)))) %>% 
  group_by(decade) %>% 
  summarize(m = mean(mPerMovie))
```
In addition, Hatch's claim does not take into account that many of the original films included in the IMDb database were small-scale in nature (unknown to the average moviegoer). The volume of movies that have been released is somewhat misleading, as it assumes that quantity overrules name recognition in terms of influence. Though this may be true in some industries, it is not true in film: movie franchises are almost exclusively produced by major studios, which have far larger reaches than small, independent companies.

	
Although Hollywood is putting out less sequels as a percentage on the whole, that does not change the fact that we perceive the amount of sequels to be increasing. In particular, increasing longevity of franchises and decreasing time between movies creates the illusion that sequels have completely taken over Hollywood. In order to change this trend, moviegoers must make conscious shifts in their mindsets. It will involve risk, but choosing the new and unknown over the predictably comforting is the only way this pattern will change any time soon.